\documentclass[numbers=endperiod, 11pt]{scrartcl}
\input{../preamble.tex}
\ihead{\textbf{Pragyan Pranay}}
\ohead{\footnotesize\url{satisfiedmagma.github.io}}

\begin{document}
\title{The Limit Superior and Limit Inferior}
\subtitle{Your Reliable Ally}
\author{Pragyan Pranay}
\date{\today}
\maketitle

\begin{abstract}
	This article aims to demistify the concept of
	$\limsup$ and $\liminf$ which students have to learn
	in an introductory analysis course right after
	learning about the rigorous definition of convergence
	in $\RR$. We will build this concept from ground up
	so the idea feels natural to the students.
\end{abstract}

\tableofcontents
\newpage

\section{Reviewing Convergence in Reals} %put \RR here
We'll start by recalling the definition of convergence
of a real sequence $a_n$ in $\RR$.

\begin{definition}[Convergence in $\RR$]
	We say that a sequence $a_n$ converges in $\RR$ if
	there exists some $a \in \RR$ such that for
	any $\varepsilon > 0$ there exists $N \in \NN$ such
	that
	\[|a_n - a| < \varepsilon \quad \text{for any $n > N.$} \]
\end{definition}

The idea for this definition is quite simple, for large
enough $N$, the distance between the terms of the sequence
$a_n$ and $a$ can get arbitrarily small. Essentially,
we are \emph{analyzing} the \textbf{long-term `settling'
	behaviour of the sequence.}

This definition has some problems. First of all, if such
an $a$ exists, then is it unique? Does it even exist? The
answer to the first problem is given by the following
theorem.

\begin{theorem}[Uniqueness]
	If a real sequence $a_n$ converges to both $a$ and $a'$,
	then $a = a'$. In other words, the limit of the sequence
	is unique.
\end{theorem}

Due to the above theorem, we can now introduce a notation.
If $a_n$ converges to $a$, then we shall write it as
$\displaystyle \lim_{n \to \infty} a_n = a$ or $a_n \to a$
as $n \to \infty$.

If you consider a sequence $a_n = (-1)^n$, then it can
be proven that $a_n$ doesn't converge to any real number.
So, given any sequence, it is not necessary that the
sequence must converge.

The majority of real analysis, deals with convergent
sequences. The sequences which don't converge to anything,
treating them becomes harder. Here's a theorem which
has extensive use throughout analysis.

\begin{theorem}[Limit Inequality]\label{thm:lim_ineq}
	Let $a_n$ and $b_n$ be two real sequences such that
	$a_n \ge b_n$ and both the sequences are
	convergent in $\RR$. Then
	\[ \lim_{n \to \infty} a_n \ge \lim_{n \to \infty} b_n \]
\end{theorem}

The theorem has a \emph{strong hypothesis;} it assumes
the convergence of both the sequences. If the sequences
are not convergent, then what shall we do? We have
now finally hit the grave the problem, if the sequence ceases
to converge, then is there any useful notion to find where
the sequence is \emph{settling} in the long run?

\subsection{A Grave Problem}
\begin{example}[A Really (Un?)Settling Sequence]\label{example:multiple_accumulation_points}
	Consider the sequence $\{a_n\}_{n \ge 1}$ given by
	\[a_n = (-1)^{n}\left( 2 + \frac 1 n \right).  \]
	Is $a_n$ convergent? Does it \emph{seem to converge}
	anywhere?
\end{example}

For the beginners, it is a good exercise to sit down and
write a formal proof that the sequence doesn't converge.
Although, if you take large $n$, then one realizes
that $a_n$ seems to sometimes approach $2$ but sometimes
approaches $-2$. Here's a small table, for values of $n$
around $1000$.
\[
	\begin{array}{c|c}
		n    & a_n             \\ \hline
		997  & -2.001003\ldots \\
		998  & 2.001002\ldots  \\
		999  & -2.001001\ldots \\
		1000 & 2.001000\ldots  \\
		1001 & -2.000999\ldots \\
		1002 & 2.000998\ldots  \\
		1003 & -2.000997\ldots
	\end{array}
\]

Formally, we can write it as
\[\lim_{n \to \infty} a_{2n} = 2 \quad \text{\&} \quad \lim_{n \to \infty} a_{2n-1} = -2.\]

This means that even though the sequence doesn't
converge(due to \Cref{thm:subseq_conv}), there is still some
settling behaviour in the sequence. To analyze the
long term `settling' behaviour, we definitely want to
capture this behaviour.

\subsection{Important Terminologies}
With the motivation set in space, we'll now introduce a new
definition.

\begin{definition}[Accumulation Points]
	For a real valued sequence $\{a_n\}_{n \ge 1}$, we
	say that $u$ is an \emph{accumulation point} if
	for any $\varepsilon > 0$ there exists infinitely
	many $m \in \NN$
	such that
	\[a_m \in (u-\varepsilon, u + \varepsilon). \]
\end{definition}

In essence, this definition says that there are infinitely
members of sequence $a_n$ which are lying arbitrarily
arbitrarily close to $u$. A similar definition which
textbooks use is the following.

\begin{definition}[Subsequential Limits]
	We say $u$ is a subsequential limit of sequence
	$\{a_n\}_{n \ge 1}$ if there exists some subsequence
	$\{a_{n_k}\}_{k \ge 1}$ such that $a_{n_k} \to u$ as
	$n \to \infty$.
\end{definition}

Its not a surprise that any accumulation point must be a
subsequential limit of the sequence. Think like this,
any subsequential limit MUST be an accumulation
point by very definition of convergence. If you have
accumulation point in hand, then first pick an element
within $2^{-1}$ neighborhood of $u$. Then pick a new
element from $2^{-2}$ neighborhood. Rinse and repeat!
This way one can construct a subsequence converging to
$u$. Details are obviously missing. Thus, we shall
formally state it and prove it as a theorem.

\begin{theorem}\label{thm:accum-pt_is_subseq_lim}
	$v \in \RR$ is an \emph{accumulation point} if and
	only if $v$ is a \emph{subsequential limit}.
\end{theorem}
\begin{proof}
	{\Large To be Added\dots}
\end{proof}

With help of above theorem, we can now use the words
\emph{accumulation points} and \emph{subsequential limits}
interchangably. The following theorem becomes quite
trivial.

\begin{theorem}
	A sequence $\{a_n\}$ is converges to $a\in \RR$ if and
	only if $\{a_n\}_{n \ge 1}$ has only a unique
	\emph{accumulation point}, $a$ to be particular.
\end{theorem}

\begin{exercise}[Optional]
	Write down formal proof to the above theorem if you
	are a beginner or look it up if you are feeling stuck.
\end{exercise}

The idea of $\limsup$ is exactly the same as finding
the ``largest'' accumulation point. Its a good time to 
revisit \Cref{example:multiple_accumulation_points}.

We could clearly clearly see that $a_n$ was kind of
settling around the points $2$ and $-2$. Indeed, due to 
\Cref{thm:accum-pt_is_subseq_lim} both $2$ and $-2$ are 
\emph{accumulation points}. It requires some work to show 
that there are no other accumulation points other than 
than $2,-2$.
\begin{exercise}
	Show that the sequence 
	\[a_n = (-1)^{n}\left( 2 + \frac 1 n \right)\]
	has only two \emph{accumulation points}, which are
	$2$ and $-2$.
\end{exercise}

The largest \emph{accumulation point} of $a_n$ is $2$, thus
we would write $\limsup a_n = 2$. There are some technical
details which one needs to address, like its existence
and whether a true maximum\footnote{if you consider a 
set of all accumulation points, it could be the 
case a \emph{supremum} exists, but a maximum doesn't.}
exists or not. The next section is dedicated to properly 
establishing this notion and filling any holes of this 
concept which we have in mind.

\section{Introducing Limit Superior}

We'll walk through a few equivalent definitions and give 
the proper proofs wherever necessary. We will start with 
the subsequential limits definition and then move onto 
the more \emph{harder} definition which allows for easier
computation of the limit superior.

\begin{center}
	{\color{red}\textbf{Disclaimer:}} All the sequences
	are bounded unless stated otherwise.
\end{center}

\subsection{Largest Accumulation Point/Subsequential Limit}
We have now seen that the limit of the sequence is indeed
an \emph{accumulation point}. The idea of limit supremum
is finding the largest \emph{accumulation point} of
your $a_n$. There is no guarantee whether a true
`maximum' exists or not. Therefore, for sequence
$a_{n}$ define

\[V \coloneqq \{x \in \RR : x\text{ is an \emph{accumulation point} of } a_n\}. \]

We want to find $\sup(V)$, for which we shall use
\nameref{axiom:completeness}. We need to ensure that
$V$ is bounded above and non-empty. The following
lemma will do it.

\begin{lemma}
	$V$ is bounded above and non-empty.
\end{lemma}
\begin{proof}
	Let $u \in \RR$ be an \emph{accumulation point} or a
	\emph{subsequential limit}. Thus, there exists some
	subsequence $\{a_{n_k}\}_{k \ge 1}$ such that 
	$a_{n_k} \to u$ as $k \to \infty$. Since $a_n$ is
	bounded, there exists some $M > 0$ such that 
	$-M \le a_n \le M$ for all $n \ge 1$. Therefore, 
	\[-M \le a_{n_k} \le M \quad \text{for all $k \ge 1$.}\]
	Taking $m \to \infty$ on both the sides with aid of
	\Cref{thm:lim_ineq}, we shall $|u| < M$ for any
	accumulation point $u$ of $a_n$, proving that 
	$V$ \textbf{is bounded}.

	Since $a_n$ is bounded, we know that via 
	\nameref{thm:bolzano_weirstrass}, there exists a
	subsequence $\{a_{m_k}\}_{k \ge 1}$ such that
	$a_{m_k} \to t$ 
	for some $t \in \RR$. Since $t$ is a \emph{subsequential
	limit}, it must be an \emph{accumulation point}, which 
	shows $t \in V \implies V \ne \emptyset$.
\end{proof}

With this preparatory lemma, we are now free to define
the following.
\begin{definition}
	For a bounded real sequence $\{a_n\}_{n \ge 1}$,
	define
	\[\limsup_{n \to \infty} a_n = \sup(V)\]
	where $V$ is set of all accumulation points of
	$\{a_n\}_{n \ge 1}$.
\end{definition}
\begin{remark*}[Notation]
	There are several notations for this quantity. People
	write it as $\varlimsup a_n$ or just $\limsup a_n$.
	We shall justify it later why we use $n \to \infty$
	below $\limsup$ in a latter section.
\end{remark*}

We have finally setup the definition of the limit 
superior. A question still remains\dots Is the supremum 
attainable? The answer turns out to be a \textbf{yes.}
The following lemma is from Rudin but is very crucial 
to get a better understanding about the `distribution' of 
\emph{accumulation points} on the real line.

If a student wants to freshen up on the concept of what a
`closed' set is, then they can refer \textbf{here} in the 
appendix.

\begin{lemma}
	The \emph{subsequential limits} of a real sequence
	$\{p_n\}_{n \ge 1}$ forms a closed subset of $\RR$.
\end{lemma}

\begin{proof}
	To be added\dots
\end{proof}

\begin{exercise}[General Metric Spaces]
	Write a general result, replacing $\RR$ by a general 
	metric space $(X,d)$ where $X$ is the set of elements
	and $d$ is the distance function. Write a formal 
	proof as well.
\end{exercise}


This definition is a little difficult to handle. Finding
accumulation points in itself takes quite some computation
and then we are here talking about the `largest one'. In
particular, we want some properties of this number which
is actually useful in practice.

Regardless, this is simplest definition to introduce to 
a student initially. Instead of directly stating another
equivalent definition, we'll take sometime analyzing what
kind of behaviour should our `largest' accumulation
point exhibit.

\subsection{The Epsilon Criteria}

In this subsection, we will carry-forward and build up
more on the previous definition of limit superior.
We wish to analyze what kind of behaviour does the 
largest \emph{accumulation point} exhibits.

\begin{definition}
	For a bounded sequence $\{a_n\}_{n \ge 1}$, let
	$s^\ast \in \RR$ such that
	\begin{itemize}
		\item For any $\varepsilon > 0$, there are
		      finitely many $n$ satisfying
		      $s^\ast + \varepsilon > a_n$;
		\item For any $\varepsilon > 0$, there are
		      infinitely many $n$ satisfying
		      $s^\ast - \varepsilon < a_n$.
	\end{itemize}

	Then there is only one $s^\ast \in \RR$ which
	satisfies both of these properties. Therefore, we 
	now define
	\[ \displaystyle s^\ast \coloneqq \varlimsup_{n \to \infty} a_n.\]
\end{definition}


\subsection{Tailed Supremums}

\subsection{The Limit Inferior}

\subsection{Useful Results}

\subsection{Heuristics: When to use it?}

\newpage
\section{Few Applications}

\subsection{Artifical Problems}

\subsection{The Ratio and Root Tests}

\subsection{Radius of Convergence of Power Series}

\subsection{An Example from Probability Theory}
This one too niche, regardless a good example. If one is 
not aware of the notion of weak convergence/convergence 
in distribution in probability theory, then feel free to
skip this one for now.

\begin{prob*}{(No Atoms in Weak Convergence)}{}
	Let $(\Omega, \mathcal F, \mathbb P)$ be a
	probability space and $\{X_n\}_{n \ge 1}$ be a sequence
	of real valued-random variables and $X$ be another
	random variable such that $X_n \overset{d}{\to} X$.
	If $a$ is a continuity point of $F_X$, then prove that
	\[\lim_{m \to \infty} \mathbb{P}(X_m = a) = \mathbb{P}(X = a) = 0.  \]
\end{prob*}

\section{Limit Superior and Limit Inferior of Sets}

\subsection{}

\section{Brief Note on Extended Reals}

\newpage
\section{Practice Problems}

\begin{problem}[First Borel-Cantelli Lemma]
	Given a probability space $(\Omega, \mathcal F, \PP)$
	and a sequence events $\{A_n\}_{n \ge 1}$ where 
	$A_m \in \mathcal F$ for all $m \in \NN$, then prove
	that 
	\[\sum_{i = 1}^\infty \PP(A_i) < \infty \implies \PP\left(\limsup_{n \to \infty} A_n\right) = 0.\]
\end{problem}

\newpage
\appendix
\section{Appendix}

\subsection{Standard Facts from Analysis-I}

\begin{axiom}[The Completeness Axiom]\label{axiom:completeness}
	For any non-empty $A \subset \RR$ bounded above, the
	supremum of $A$ or $\sup(A)$ exists in $\RR$.
\end{axiom}

\begin{theorem}[Subsequences of Convergence Sequences]\label{thm:subseq_conv}
	If $a_n \to a$ as $n \to \infty$, then any
	subsequence of $a_n$ must converge to $a$ as well.
\end{theorem}

\begin{theorem}[Bolzano Weirstrass]\label{thm:bolzano_weirstrass}
	For any bounded sequence $\{b_n\}_{n \ge 1}$, there 
	exists a convergent subsequence
	$\{b_{n_k}\}_{k \ge 1}$.
\end{theorem}


\end{document}
